---
title: "Massachusetts Bridge Condition Report"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)

bridges <- read_csv("data/MA_bridges_2025.csv", show_col_types = FALSE)

county_names <- c(
  "001" = "Barnstable", "003" = "Berkshire",  "005" = "Bristol",
  "007" = "Dukes",      "009" = "Essex",       "011" = "Franklin",
  "013" = "Hampden",    "015" = "Hampshire",   "017" = "Middlesex",
  "019" = "Nantucket",  "021" = "Norfolk",     "023" = "Plymouth",
  "025" = "Suffolk",    "027" = "Worcester"
)

cond_labels <- c(
  "0" = "0 – Failed",
  "1" = "1 – Imminent Failure",
  "2" = "2 – Critical",
  "3" = "3 – Serious",
  "4" = "4 – Poor",
  "5" = "5 – Fair",
  "6" = "6 – Satisfactory",
  "7" = "7 – Good",
  "8" = "8 – Very Good",
  "9" = "9 – Excellent"
)

bridges <- bridges |>
  mutate(
    county = county_names[COUNTY_CODE_003],
    DECK_COND_058 = as.character(DECK_COND_058),
    SUPERSTRUCTURE_COND_059 = as.character(SUPERSTRUCTURE_COND_059),
    SUBSTRUCTURE_COND_060 = as.character(SUBSTRUCTURE_COND_060)
  )

condition_colors <- c("G" = "#2ecc71", "F" = "#f39c12", "P" = "#e74c3c")
rating_palette <- c(
  "#8B0000", "#e74c3c", "#e74c3c", "#e74c3c", "#e74c3c",
  "#f39c12", "#f39c12", "#2ecc71", "#2ecc71", "#27ae60"
)
```

## Overview

This report summarizes the structural condition of **`r format(nrow(bridges), big.mark=",")`
bridges** in Massachusetts, based on 2025 National Bridge Inventory (NBI) data. Each bridge
is rated on three components — deck, superstructure, and substructure — using a 0–9 scale.
The overall `BRIDGE_CONDITION` classification (Good / Fair / Poor) is derived from the
lowest of those three component ratings.

```{r}
#| label: fig-overall
#| fig-cap: "Overall bridge condition across Massachusetts."

overall <- bridges |>
  count(BRIDGE_CONDITION) |>
  mutate(
    label = case_when(
      BRIDGE_CONDITION == "G" ~ "Good",
      BRIDGE_CONDITION == "F" ~ "Fair",
      BRIDGE_CONDITION == "P" ~ "Poor"
    ),
    label = factor(label, levels = c("Good", "Fair", "Poor")),
    pct = n / sum(n)
  )

ggplot(overall, aes(label, n, fill = BRIDGE_CONDITION)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%s\n%.1f%%", format(n, big.mark=","), 100*pct)),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = condition_colors, guide = "none") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Number of bridges") +
  theme_minimal(base_size = 13)
```

::: {.callout-warning}
**`r bridges |> filter(BRIDGE_CONDITION == "P") |> nrow()` bridges** (`r sprintf("%.1f%%", 100 * mean(bridges$BRIDGE_CONDITION == "P"))` of total) are rated **Poor** — meaning at least one structural component has advanced deterioration (rating ≤ 4). These bridges may carry load restrictions or require monitoring.
:::

## Condition Ratings

Bridges are inspected on three structural components. A rating of **5 (Fair)** means
primary elements are sound but show minor cracking or section loss. Ratings of **4 (Poor)**
or below indicate significant deterioration.

```{r}
#| label: fig-components
#| fig-cap: "Rating distribution for each of the three structural components. Bridges rated N (not applicable, e.g. culverts) are excluded."
#| fig-height: 7

component_data <- bind_rows(
  bridges |> filter(DECK_COND_058 != "N") |>
    count(rating = DECK_COND_058) |> mutate(component = "Deck"),
  bridges |> filter(SUPERSTRUCTURE_COND_059 != "N") |>
    count(rating = SUPERSTRUCTURE_COND_059) |> mutate(component = "Superstructure"),
  bridges |> filter(SUBSTRUCTURE_COND_060 != "N") |>
    count(rating = SUBSTRUCTURE_COND_060) |> mutate(component = "Substructure")
) |>
  mutate(
    rating = factor(rating, levels = as.character(0:9)),
    fill_group = case_when(
      rating %in% c("0","1","2","3","4") ~ "Poor",
      rating %in% c("5","6")             ~ "Fair",
      TRUE                               ~ "Good"
    ),
    fill_group = factor(fill_group, levels = c("Good","Fair","Poor"))
  )

ggplot(component_data, aes(rating, n, fill = fill_group)) +
  geom_col() +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c("Good"="#2ecc71","Fair"="#f39c12","Poor"="#e74c3c")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Condition rating", y = "Number of bridges", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-component-summary
#| tbl-cap: "Percentage of bridges in each condition band by component (N-rated excluded)."

component_summary <- tibble(
  Component = c("Deck", "Superstructure", "Substructure"),
  Poor = c(
    bridges |> filter(DECK_COND_058 != "N", as.integer(DECK_COND_058) <= 4) |> nrow(),
    bridges |> filter(SUPERSTRUCTURE_COND_059 != "N", as.integer(SUPERSTRUCTURE_COND_059) <= 4) |> nrow(),
    bridges |> filter(SUBSTRUCTURE_COND_060 != "N", as.integer(SUBSTRUCTURE_COND_060) <= 4) |> nrow()
  ),
  Fair = c(
    bridges |> filter(DECK_COND_058 %in% c("5","6")) |> nrow(),
    bridges |> filter(SUPERSTRUCTURE_COND_059 %in% c("5","6")) |> nrow(),
    bridges |> filter(SUBSTRUCTURE_COND_060 %in% c("5","6")) |> nrow()
  ),
  Good = c(
    bridges |> filter(DECK_COND_058 %in% c("7","8","9")) |> nrow(),
    bridges |> filter(SUPERSTRUCTURE_COND_059 %in% c("7","8","9")) |> nrow(),
    bridges |> filter(SUBSTRUCTURE_COND_060 %in% c("7","8","9")) |> nrow()
  ),
  Total = c(
    bridges |> filter(DECK_COND_058 != "N") |> nrow(),
    bridges |> filter(SUPERSTRUCTURE_COND_059 != "N") |> nrow(),
    bridges |> filter(SUBSTRUCTURE_COND_060 != "N") |> nrow()
  )
) |>
  mutate(
    `% Poor` = sprintf("%.1f%%", 100 * Poor / Total),
    `% Fair`  = sprintf("%.1f%%", 100 * Fair  / Total),
    `% Good`  = sprintf("%.1f%%", 100 * Good  / Total)
  ) |>
  select(Component, Poor, `% Poor`, Fair, `% Fair`, Good, `% Good`)

knitr::kable(component_summary, align = "lrrrrrr")
```

## Condition by County

```{r}
#| label: fig-county
#| fig-cap: "Bridge condition by county, ordered by percentage rated Poor."
#| fig-height: 6

county_cond <- bridges |>
  filter(!is.na(county)) |>
  count(county, BRIDGE_CONDITION) |>
  group_by(county) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  mutate(
    label = case_when(
      BRIDGE_CONDITION == "G" ~ "Good",
      BRIDGE_CONDITION == "F" ~ "Fair",
      BRIDGE_CONDITION == "P" ~ "Poor"
    ),
    label = factor(label, levels = c("Good","Fair","Poor"))
  )

county_order <- county_cond |>
  filter(BRIDGE_CONDITION == "P") |>
  arrange(pct) |>
  pull(county)

county_cond |>
  mutate(county = factor(county, levels = county_order)) |>
  ggplot(aes(pct, county, fill = label)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Good"="#2ecc71","Fair"="#f39c12","Poor"="#e74c3c")) +
  labs(x = "Share of bridges", y = NULL, fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

Essex County has the highest share of Poor-rated bridges at **16.6%**, followed by
Franklin (12.4%) and Berkshire (11.6%). More rural counties with older infrastructure
tend to have higher rates of deterioration.

## Age and Condition

The median Massachusetts bridge was built in **`r median(bridges$YEAR_BUILT_027, na.rm=TRUE)`**,
with a quarter of the inventory dating from before **`r quantile(bridges$YEAR_BUILT_027, 0.25, na.rm=TRUE)`**.
Bridges built in the 1920s and 1930s have the highest Poor rates.

```{r}
#| label: fig-decade
#| fig-cap: "Percentage of bridges rated Poor, by decade of construction (1910–2020s)."
#| fig-height: 4

decade_data <- bridges |>
  mutate(decade = floor(YEAR_BUILT_027 / 10) * 10) |>
  filter(decade >= 1910, !is.na(BRIDGE_CONDITION)) |>
  group_by(decade) |>
  summarise(
    n = n(),
    pct_poor = mean(BRIDGE_CONDITION == "P"),
    pct_fair = mean(BRIDGE_CONDITION == "F"),
    pct_good = mean(BRIDGE_CONDITION == "G")
  ) |>
  pivot_longer(starts_with("pct_"), names_to = "condition", values_to = "pct") |>
  mutate(
    condition = case_when(
      condition == "pct_good" ~ "Good",
      condition == "pct_fair" ~ "Fair",
      condition == "pct_poor" ~ "Poor"
    ),
    condition = factor(condition, levels = c("Good","Fair","Poor"))
  )

ggplot(decade_data, aes(decade, pct, fill = condition)) +
  geom_col(position = "stack", width = 8) +
  scale_x_continuous(breaks = seq(1910, 2020, 10)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Good"="#2ecc71","Fair"="#f39c12","Poor"="#e74c3c")) +
  labs(x = "Decade built", y = "Share of bridges", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## High-Priority Concerns

The 15 Poor-rated bridges carrying the most daily traffic represent an urgent maintenance
priority, as failures would affect large numbers of commuters.

```{r}
#| label: tbl-high-traffic
#| tbl-cap: "Top 15 Poor-rated bridges by average daily traffic (ADT)."

bridges |>
  filter(BRIDGE_CONDITION == "P") |>
  arrange(desc(ADT_029)) |>
  slice_head(n = 15) |>
  mutate(
    Route = str_trim(str_remove_all(FACILITY_CARRIED_007, "'")),
    Crosses = str_trim(str_remove_all(FEATURES_DESC_006A, "'")),
    ADT = format(ADT_029, big.mark = ","),
    Rating = LOWEST_RATING,
    Built = YEAR_BUILT_027
  ) |>
  select(Route, Crosses, County = county, Built, ADT, Rating) |>
  knitr::kable(align = "llllrr")
```

::: {.callout-note}
**`r bridges |> filter(LOWEST_RATING <= 3) |> nrow()` bridges** have a component rating of 3 (Serious) or below, indicating local failures are possible or the bridge is at imminent risk. These require the most urgent attention.
:::

## Summary

| Metric | Value |
|---|---|
| Total bridges | `r format(nrow(bridges), big.mark=",")` |
| Good condition | `r bridges |> filter(BRIDGE_CONDITION=="G") |> nrow() |> format(big.mark=",")` (`r sprintf("%.1f%%", 100*mean(bridges$BRIDGE_CONDITION=="G"))`) |
| Fair condition | `r bridges |> filter(BRIDGE_CONDITION=="F") |> nrow() |> format(big.mark=",")` (`r sprintf("%.1f%%", 100*mean(bridges$BRIDGE_CONDITION=="F"))`) |
| Poor condition | `r bridges |> filter(BRIDGE_CONDITION=="P") |> nrow() |> format(big.mark=",")` (`r sprintf("%.1f%%", 100*mean(bridges$BRIDGE_CONDITION=="P"))`) |
| Serious or worse (rating ≤ 3) | `r bridges |> filter(LOWEST_RATING <= 3) |> nrow()` |
| Median year built | `r median(bridges$YEAR_BUILT_027, na.rm=TRUE)` |
| County with most Poor bridges | Essex (16.6%) |

*Data source: Federal Highway Administration National Bridge Inventory, 2025.*
